---
title: "Growth Mixture Models"
author: "Biostatistics Working Group"

---

## Overview
Growth mixture modeling (GMM) is a statistical technique used to identify heterogeneous subgroups or classes that follow distinct developmental paths (growth trajectories) within populations over time. These models are an extension of latent growth curve models, with a focus on delineating variability in growth trajectories. This is particularly relevant for longitudinal research where it's suspected that different subpopulations might exhibit different growth, stability, or decline trajectories. The GMM framework estimates both the shape and number of latent trajectories and the probability of an individual belonging to each trajectory. These models capture overall growth patterns while simultaneously allowing for subgroup variations.

[+add diagrams/figures]

### When to use Growth Mixture Models?
You should consider Growth Mixture Models in the following situations:

1. You are interested in: Identifying subgroups or trajectories within a population that exhibit distinct growth patterns over time.
2. Your data: Captures repeated measurements, tracking the development or change of a certain variable across multiple time points.
3. Your objective: Is to uncover latent heterogeneity in growth trajectories and classify individuals into these distinct trajectories based on observed data.

### Getting Started with Growth Mixture Models
In this tutorial, we will guide you through an example of a growth mixture models using a small dataset. By the time you complete this tutorial, you will be equipped to:

1. Understand the basic concepts behind Growth Mixture Models.
2. Implement this modeling technique on your longitudinal data using the appropriate statistical tools.
3. Analyze and interpret the outcomes to discern the distinct growth trajectories and the proportion of individuals in each trajectory.

## Basic Example
In this tutorial, we will begin by generating a sample dataset directly within R to ensure everyone has the same starting point and can follow along without needing to download or access external files. This generated dataset will serve as our example throughout this tutorial. In your own analyses, you'll likely start by importing your own data.

The simulated dataset created for this example consists of scores on a single outcome variable ("Job_Satisfaction") for 500 individuals each measured at four time points: T1-T4.  

#### Create Example Dataset
```{r}

# Setting the seed for reproducibility
set.seed(123)

# Setting the number of individuals for our sample data
n <- 500

# Specifying the desired correlation matrix for our data (5x5 matrix for 5 timepoints)
cor_matrix <- matrix(c(1, 0.3, 0.3, 0.3, 0.3,
                      0.3, 1, 0.3, 0.3, 0.3,
                      0.3, 0.3, 1, 0.3, 0.3,
                      0.3, 0.3, 0.3, 1, 0.3,
                      0.3, 0.3, 0.3, 0.3, 1), 
                      nrow=5)

# Using Cholesky decomposition to generate correlated data
cholesky <- chol(cor_matrix)

# Generating correlated random data for 5 timepoints
random_data <- matrix(rnorm(n * 5), ncol=5)
correlated_data <- random_data %*% cholesky

# Converting the generated data into a structured data frame in wide format
data_wide <- data.frame(
  Individual = 1:n,
  Job_Satisfaction_T1 = round(runif(n, 5, 10)),
  Job_Satisfaction_T2 = round(runif(n, 5, 10)),
  Job_Satisfaction_T3 = round(runif(n, 5, 10)),
  Job_Satisfaction_T4 = round(runif(n, 5, 10)),
  Job_Satisfaction_T5 = round(runif(n, 5, 10)),
  Life_Satisfaction_T1 = round(correlated_data[,1] * 2 + 7),
  Life_Satisfaction_T2 = round(correlated_data[,2] * 2 + 7),
  Life_Satisfaction_T3 = round(correlated_data[,3] * 2 + 7),
  Life_Satisfaction_T4 = round(correlated_data[,4] * 2 + 7),
  Life_Satisfaction_T5 = round(correlated_data[,5] * 2 + 7)
)

```

### Model Specification and Estimation
Now that we have our simulated data, we can fit a Growth Mixture Model. In this example, we will assume that there are three latent classes. However, in practice, the number of classes is usually unknown and must be determined through model comparisons or other methods.

### Model Specification and Estimation
To specify a Growth Mixture Model (GMM), we will leverage the lcmm syntax to identify distinct trajectory classes within the data, representing different patterns of growth or change. This enables us to capture the heterogeneity in developmental trajectories, considering both the shared and individual-specific growth patterns.

#### Install and Load Necessary Libraries
```{r setup, echo=TRUE, warning=FALSE, message=FALSE}
# Install the lcmm package if not already installed
if (!("lcmm" %in% installed.packages())) {
  install.packages("lcmm")
}

library(lcmm)

```

#### Model
```{r}

# Load necessary libraries
library(lcmm)
library(tidyr)

# Convert wide format data to long format
data_long <- data_wide %>% 
  gather(key = "Variable_Time", value = "Score", -Individual) %>%
  separate(Variable_Time, into = c("Variable", "Time"), sep = "_T")

# Specify initial values
Bfixed <- c(7, 0.5, 7, 0.5) # Initial values for intercept and slope for each class
Brand <- c(7, 0.5, 4, 0.2, 0.1)  # Initial values for random intercept, slope, their variances, and covariance
Bmix <- c(7, 0.5, 7, 0.5) # Initial values for mixture intercept and slope for each class

# Fit the GMM for Life Satisfaction across 5 time points
gmm_model <- lcmm(Score ~ Time, random = ~ Time, subject = 'Individual', 
                 mixture = ~ Time, 
                 ng = 2, # specifying number of latent classes 
                 B = list(fixed = Bfixed, random = Brand, mixture = Bmix),
                 data = subset(data_long, Variable == "Life_Satisfaction"))

# Examine the results
summary(gmm_model)

```

#### Fit Model
```{r}
# Obtain and print a summary of the model
summary_gmm <- summary(gmm)
```

#### Model Summary
```{r}
# Print a summary of the model
print(summary_gmm)
```

### Interpreting the Results
A Growth Mixture Model (GMM) was applied to identify distinct trajectories (or growth patterns) in the dataset. GMM allows for the identification of latent subgroups within a population that follow different developmental or growth trajectories over time.

Using the `hlme` function, we estimated a GMM with 3 distinct growth trajectories (subgroups) using the predictor `time`. 

From the `summary` output:
- The estimated means and variances for each trajectory group can be extracted. 
- The proportions of individuals in each trajectory group are: `r gmm$proportions`
- The estimated growth parameters (intercept and slope) for each trajectory group are: `r gmm$parameters`

## Conclusion
The Growth Mixture Model has identified multiple distinct trajectories within the data. By examining the estimated proportions and growth parameters, we can gain insights into the different developmental or growth patterns present in the population. Further, statistical tests can be conducted to determine the significance and implications of these trajectories.






