---
title: "Multivariate Latent Growth Curves"
author: "Biostatistics Working Group"
---

## Overview
Multivariate latent growth curve models (MLGCMs) are a statistical technique that enables the examination of multiple growth trajectories simultaneously. These models are particularly useful for longitudinal investigations in which several related variables or outcomes develop concurrently, and the interplay between them is of interest. This approach uses a structural equation modeling framework to elucidate both the individual trajectories (fixed effects) and the variability around these trajectories (random effects) for multiple correlated growth processes. This simultaneous exploration of multiple "latent" growth curves, each capturing a distinct yet related developmental trajectory, is the defining feature of these models.

[+add diagrams/figures]

### When to use Multivariate Latent Growth Curves?
You should consider multivariate latent growth curves in the following situations:

1. You are interested in: Examining the parallel trajectories of multiple measured variables over time.
2. Your data: Incorporates repeated measures for multiple variables at several time points.
3. You wish to: Understand the interplay between these variables as they change.

### Getting Started with Multivariate Latent Growth Curves
In this tutorial, we will guide you through an example of a multivariate latent growth curve model using a small dataset. By the time you complete this tutorial, you will be equipped to:

1. Understand the basic principles of multivariate latent growth curves.
2. Implement these models on your own longitudinal data using appropriate statistical tools.
3. Analyze the outputs and make informed decisions based on the multivariate growth trajectories.

## Basic Example
In this tutorial, we will begin by generating a sample dataset directly within R to ensure everyone has the same starting point and can follow along without needing to download or access external files. This generated dataset will serve as our example throughout this tutorial. In your own analyses, you'll likely start by importing your own data.

The simulated dataset created for this example consists of scores on two outcome variables ("Job_Satisfaction", "Life_Satisfaction") for 500 individuals each measured at four time points: T1-T4. 

#### Create Example Dataset
```{r}

# Setting the seed for reproducibility
set.seed(123)

# Setting the number of individuals for our sample data
n <- 500

# Specifying the desired correlation matrix for our data
cor_matrix <- matrix(c(1, 0.3, 0.3, 0.3, 
                      0.3, 1, 0.3, 0.3,
                      0.3, 0.3, 1, 0.3,
                      0.3, 0.3, 0.3, 1), 
                      nrow=4)

# Using Cholesky decomposition to generate correlated data
cholesky <- chol(cor_matrix)

# Generating correlated random data
random_data <- matrix(rnorm(n * 4), ncol=4)
correlated_data <- random_data %*% cholesky

# Converting the generated data into a structured data frame in wide format
data <- data.frame(
  Individual = 1:n,
  Job_Satisfaction_T1 = round(runif(n, 5, 10)),
  Job_Satisfaction_T2 = round(runif(n, 5, 10)),
  Job_Satisfaction_T3 = round(runif(n, 5, 10)),
  Job_Satisfaction_T4 = round(runif(n, 5, 10)),
  Life_Satisfaction_T1 = round(correlated_data[,1] * 2 + 7),
  Life_Satisfaction_T2 = round(correlated_data[,2] * 2 + 7),
  Life_Satisfaction_T3 = round(correlated_data[,3] * 2 + 7),
  Life_Satisfaction_T4 = round(correlated_data[,4] * 2 + 7)
)

```

### View Dataset
```{r}
head(data)
```

### Model Specification and Estimation
To specify a Multivariate Latent Growth Curve Model (MLGCM), we will utilize the lavaan syntax to outline both the measurement model and the structural model. In an MLGCM, we simultaneously model the trajectories of multiple outcome variables across various time points. This approach allows us to examine the parallel growth patterns of these variables and to explore their interrelationships, capturing shared growth dynamics and distinct individual trajectories.

#### Install and Load Necessary Libraries
```{r}
# Load necessary packages
if (!("lavaan" %in% installed.packages())) install.packages("lavaan")
library(lavaan)
```

#### Model
```{r}

# Specify the multivariate latent growth curve model
model <- '
  # Intercept and slope factors for variable 1
  i_var1 =~ 1*Job_Satisfaction_T1 + 1*Job_Satisfaction_T2 + 1*Job_Satisfaction_T3
  s_var1 =~ 0*Job_Satisfaction_T1 + 1*Job_Satisfaction_T2 + 2*Job_Satisfaction_T3

  # Intercept and slope factors for variable 2
  i_var2 =~ 1*Life_Satisfaction_T1 + 1*Life_Satisfaction_T2 + 1*Life_Satisfaction_T3
  s_var2 =~ 0*Life_Satisfaction_T1 + 1*Life_Satisfaction_T2 + 2*Life_Satisfaction_T3
'
```

#### Fit Model
```{r}
# Estimate the model
fit <- lavaan::sem(model, data = data)
```

#### Model Summary
```{r}
summary(fit, fit.measures=TRUE, standardized=TRUE, rsquare=TRUE)
```

```{r}

## Creating variable (names) from output to be used for inline text

library(broom)
tidyfit <- tidy(fit)
#augmentfit <- augment(fit)
glancefit <- glance(fit)

# Extract column data
#p_values <- tidyfit$p.value
#estimate <- tidyfit$estimate

# Extract p-values for specific terms into tibbles
intercept_variance_estimate_V1 <- round(tidyfit[tidyfit$term == "i_var1 ~~ i_var1", "estimate"], 3)
slope_variance_estimate_V1 <- round(tidyfit[tidyfit$term == "s_var1 ~~ s_var1", "estimate"], 3)

intercept_variance_estimate_V2 <- round(tidyfit[tidyfit$term == "i_var2 ~~ i_var2", "estimate"], 3)
slope_variance_estimate_V2 <- round(tidyfit[tidyfit$term == "s_var2 ~~ s_var2", "estimate"], 3)

rmsea <- glancefit$rmsea
bic <- glancefit$BIC

# Extract values from the tibble for inline text
var_i_V1 <- intercept_variance_estimate_V1$estimate
var_s_V1 <- slope_variance_estimate_V1$estimate

var_i_V2 <- intercept_variance_estimate_V2$estimate
var_s_V2 <- slope_variance_estimate_V2$estimate

#model_list <- split(tidyfit, tidyfit$term)

```

### Interpreting the Results
A Multivariate Latent Growth Curve Model (MLGCM) was used to understand the trajectories of two variables (`Job Satisfaction` and `Life Satisfaction`) over time. This model creates separate growth curves for each variable and allows for an examination of how the trajectories of the two variables relate to each other.

For each variable, the model separates the growth into two latent factors:
- `Intercept (i)`: Represents the initial status or starting point.
- `Slope (s)`: Represents the rate of change over time.

From the `sem` output:
- For `Job Satisfaction`, the average starting point (intercept) is: `r var_i_V1` and the average rate of change (slope) is: `r var_s_V1`
- For `Life Satisfaction`, the average starting point (intercept) is: `r var_i_V2` and the average rate of change (slope) is: `r var_s_V2`

The RMSEA of the model is `r rmsea`.
The BIC of the model is `r round(bic, 3)`.

## Conclusion
The MLGCM provides insights into how the trajectories of `Job Satisfaction` and `Life Satisfaction` evolve over time. By examining the estimated intercepts and slopes for both variables, we can understand their growth patterns and how they might be related. Further, covariances between the latent factors can provide information on whether the initial status or rate of change in one variable is related to that of the other variable.

